---
title: "PHB_221_Final_Project"
author: "Abigail, Jade, Yinan"
date: "2025-11-13"
output: pdf_document
---

# Main Questions
Main Question: How does the age-adjusted death rate for heart disease vary across states in the United States, and are there any observable regional or geographic trends over time?

- Research Objectives:
  - Compare state-level heart disease mortality rates
  - Identify patterns
  - Explore how rates change over time
  - Determine whether areas with consistently higher mortality are geographically concentrated 
  - Can past trends in age adjusted heart disease death rates be used to predict future state level rates?

# Dataset and Libraries

```{r}
# libraries
library(tidyverse)
library(lme4)
library(Metrics)
# dataset
data <- read_csv("NCHS_-_Leading_Causes_of_Death__United_States.csv")
```

# Data Collection and Cleaning

```{r}
# view first few rows of data
head(data)

# check the number of rows and cols
dim(data)

# filter data
filtered_data <- data[data$`Cause Name` == "Heart disease", ]

head(filtered_data)

dim(filtered_data)

# missing data
colSums(is.na(filtered_data))
```

# Explantory Analysis

```{r}

# national level
national_data <- data |>
  filter(`Cause Name` == "Heart disease", State == "United States")

head(national_data)

ggplot(data = national_data, mapping = aes(x = Year, y = `Age-adjusted Death Rate`)) +
  geom_line() + 
  labs(title = "National Age Adjusted Heart Disease Rate Over Time")

# state level

state_only <- filtered_data |>
  filter(State != "United States") |>
  mutate(
    Region = case_when(
      State %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts",
                   "Rhode Island", "Connecticut", "New York", "New Jersey",
                   "Pennsylvania") ~ "Northeast",
      State %in% c("Ohio", "Indiana", "Illinois", "Michigan", "Wisconsin",
                   "Minnesota", "Iowa", "Missouri", "North Dakota", "South Dakota",
                   "Nebraska", "Kansas") ~ "Midwest",
      State %in% c("Delaware", "Maryland", "District of Columbia", "Virginia",
                   "West Virginia", "North Carolina", "South Carolina", "Georgia",
                   "Florida", "Kentucky", "Tennessee", "Alabama", "Mississippi",
                   "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
      State %in% c("Montana", "Idaho", "Wyoming", "Colorado", "New Mexico",
                   "Arizona", "Utah", "Nevada", "Washington", "Oregon", "California",
                   "Alaska", "Hawaii") ~ "West"
    )
  )


head(state_only)

ggplot(data = state_only, mapping = aes(x = Year, y = `Age-adjusted Death Rate`)) + 
  geom_line() + 
  facet_wrap(~ State) +
  labs(title = "State Age Adjusted Heart Disease Rate Over Time")


# regional level
regional_summary <- state_only |>
  group_by(Region, Year) |>
  summarise(
    Rate = mean(`Age-adjusted Death Rate`)
  )


ggplot(data = regional_summary, mapping = aes(x = Year, y = Rate, color = Region)) + 
  geom_line(linewidth = 1) +
  labs(title = "Regional Age Adjusted Heart Disease Rate Over Time")

ggplot(data = state_only, mapping = aes(x = Region, y = `Age-adjusted Death Rate`, color = Region)) +
  geom_boxplot(staplewidth = 0.5) +
  labs(title = "Age-Adjusted Heart Disease Mortality by Region")

```


# Statistical Testing

- How does mortality vary across states, and are there regional/geographical patterns over time?

- one way anova limitations:
  - has to be done using a single year, not all years
  - anova assumes independent groups
  - using all years violate independence

- two way anova


- notes:
  - https://www.geeksforgeeks.org/data-science/analysis-of-variance-anova/
  - cook's distance
    - How does a single observation (data pt) influece the entire model?
    - If a data pt is removed, does the model's prediction change a lot?
  
```{r}
# two way anova testing
# * (interaction term) bc each state has their own slope, if + was used then slope for Year is the same across all states
two_way <- aov(`Age-adjusted Death Rate` ~ State * Year, data = state_only)
summary(two_way)

# check assumptions: homoscedasticity, normalit of residuals, linearity, and influential pts
par(mfrow = c(2,2))
plot(two_way)

D <- cooks.distance(two_way)
sum(D > 1)


```



# Predictive Modeling

- Can past trends in age adjusted heart disease death rates be used to predict future state level rates?

- notes:
  - simple linear regression wont work bc we need year and state 
    - assumes all states share the same mean and trend over time
  
  - multiple linear regression year + state
    - treats state as a categorical predictor (fixed state effects)
    - each state getes it's own intercept (baseline level)
    - BUT it assumes same slope for all states
    
  - multiple linear regression year * state 
    - different slope per state

```{r}
# simple linear regression
m1 <- lm(`Age-adjusted Death Rate` ~ Year, data = state_only)
#summary(m1)


# multiple regression +
m2 <- m_additive <- lm(`Age-adjusted Death Rate` ~ Year + State, data = state_only)
#summary(m2)

# multiple regression *
m3 <- lm(`Age-adjusted Death Rate` ~ Year * State, data = state_only)
#summary(m3)

# mixed model?

mixed1 <- lmer(`Age-adjusted Death Rate` ~ Year + (1 | State), data = state_only)
#summary(mixed1)

mixed2 <- lmer(`Age-adjusted Death Rate` ~ Year + (Year | State), data = state_only)
#summary(mixed2)


# plot comparison

state_only <- state_only |>
  mutate(
    fitted_m1 = fitted(m1),
    fitted_m2 = fitted(m2),
    fitted_m3 = fitted(m3),
    fitted_mixed1 = fitted(mixed1),
    fitted_mixed2 = fitted(mixed2)
  )

head(state_only)


ggplot(data = state_only, mapping = aes(x = Year)) +
  
  # actual data 
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  # year + state
  geom_line(mapping = aes(y = fitted_m2, color = "Fitted: Year + State"), linetype = "dashed") +
  
  # year * state
  geom_line(mapping = aes(y = fitted_m3, color = "Fitted: Year * State"), linetype = "dotted") +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(
    title = "Comparison of (Year + State) vs Interaction (Year * State) Models",
    y = "Age-Adjusted Heart Disease Death Rate",
    color = ""
  ) +
  
  scale_color_manual(values = c(
    "Actual" = "black",
    "Fitted: Year + State" = "blue",
    "Fitted: Year * State" = "red"
  ))


ggplot(data = state_only, mapping = aes(x = Year)) +
  # actual data
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  # m3
  geom_line(aes(y = fitted_m3, color = "Fixed: Year * State"), linetype = "dashed") +
  
  # mixed1
  geom_line(aes(y = fitted_mixed1, color = "Mixed: (1 | State)"), linetype = "dashed") +
  
  # mixed2
  geom_line(mapping = aes(y = fitted_mixed2, color = "Mixed: (Year | State)"), linetype = "dotted") +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(
    title = "Actual vs Fixed-Effects and Mixed-Effects Models by State",
    y = "Age-Adjusted Heart Disease Death Rate",
    color = ""
  ) +
  
  scale_color_manual(values = c(
    "Actual"              = "black",
    "Fixed: Year * State" = "blue",
    "Mixed: (1 | State)"  = "darkgreen",
    "Mixed: (Year | State)" = "red"
  ))

```

- Train Models:
  - 80/10/10 does not work bc we cant randomly shuffle years

```{r}
train <- state_only |>
  filter(Year <= 2013)

test <- state_only |>
  filter(Year > 2013)

# train models
m2_train <- lm(`Age-adjusted Death Rate` ~ Year + State, data = train)
m3_train <- lm(`Age-adjusted Death Rate` ~ Year * State, data = train)
mixed1_train <- lmer(`Age-adjusted Death Rate` ~ Year + (1 | State), data = train)
mixed2_train <- lmer(`Age-adjusted Death Rate` ~ Year + (Year | State), data = train)

# predict on test set
test$pred_m2 <- predict(m2_train, newdata = test)
test$pred_m3 <- predict(m3_train, newdata = test)
test$pred_mixed1 <- predict(mixed1_train, newdata = test)
test$pred_mixed2 <- predict(mixed2_train, newdata = test)

ggplot(data = test, mapping = aes(x = Year)) +
  
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  geom_line(mapping = aes(y = pred_m3, color = "Predicted (m3)")) +
  
  geom_line(mapping = aes(y = pred_mixed1, color = "Predicted (m1)")) +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(title = "Actual vs Predicted Heart Disease Mortality (Test Set: 2014–2017)") +
  
  scale_color_manual(values = c("Actual" = "black", "Predicted (m3)" = "blue", "Predicted (mixed m1)" = "red"))


```


# Evaluation and Interpretation

- look at R^2, p vals, residuals

```{r}
# check models accuracies
# RMSE and MAE
rmse_m2 <- rmse(test$`Age-adjusted Death Rate`, test$pred_m2)
mae_m2  <- mae(test$`Age-adjusted Death Rate`, test$pred_m2)

rmse_m3 <- rmse(test$`Age-adjusted Death Rate`, test$pred_m3)
mae_m3  <- mae(test$`Age-adjusted Death Rate`, test$pred_m3)

rmse_mixed1 <- rmse(test$`Age-adjusted Death Rate`, test$pred_mixed1)
mae_mixed1  <- mae(test$`Age-adjusted Death Rate`, test$pred_mixed1)

rmse_mixed2 <- rmse(test$`Age-adjusted Death Rate`, test$pred_mixed2)
mae_mixed2  <- mae(test$`Age-adjusted Death Rate`, test$pred_mixed2)

# accuracies
data.frame(
  Model = c("m2 (Year+State)", "m3 (Year*State)", 
            "mixed1 (1|State)", "mixed2 (Year|State)"),
  RMSE = c(rmse_m2, rmse_m3, rmse_mixed1, rmse_mixed2),
  MAE  = c(mae_m2, mae_m3, mae_mixed1, mae_mixed2)
)

```
Between the two models, the predictions follow a smooth linear decline because the models learn long term state specific trends but it can't capture short term year to year flucatuatons. Based on the RMSE and MAE Year x State (m3) performed the best as it showed the lowest scores, whicbh means allowing each state to have its own slope improves predictive model performance. Mixed effects were a bit less accurate because it tries to make the slopes to be more similar across states. 

