---
title: "PHB_221_Final_Project"
author: "Abigail, Jade, Yinan"
date: "2025-11-13"
output: pdf_document
---

# Main Questions
Main Question: How does the age-adjusted death rate for heart disease vary across states in the United States, and are there any observable regional or geographic trends over time?

- Research Objectives:
  - Compare state-level heart disease mortality rates
  - Identify patterns
  - Explore how rates change over time
  - Determine whether areas with consistently higher mortality are geographically concentrated 
  - Can past trends in age adjusted heart disease death rates be used to predict future state level rates?

# Dataset and Libraries

```{r}
# libraries
library(tidyverse)
library(lme4)
library(Metrics)
library(randomForest)
library(alr4)
# dataset
data <- read_csv("NCHS_-_Leading_Causes_of_Death__United_States.csv")
```

# Data Collection and Cleaning

```{r}
# view first few rows of data
head(data)

# check the number of rows and cols
dim(data)

# filter data
filtered_data <- data[data$`Cause Name` == "Heart disease", ]

head(filtered_data)

dim(filtered_data)

# missing data
colSums(is.na(filtered_data))
```

# Explantory Analysis

```{r}

# national level
national_data <- data |>
  filter(`Cause Name` == "Heart disease", State == "United States")

head(national_data)

ggplot(data = national_data, mapping = aes(x = Year, y = `Age-adjusted Death Rate`)) +
  geom_line() + 
  labs(title = "National Age Adjusted Heart Disease Rate Over Time")

# state level

state_only <- filtered_data |>
  filter(State != "United States") |>
  mutate(
    Region = case_when(
      State %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts",
                   "Rhode Island", "Connecticut", "New York", "New Jersey",
                   "Pennsylvania") ~ "Northeast",
      State %in% c("Ohio", "Indiana", "Illinois", "Michigan", "Wisconsin",
                   "Minnesota", "Iowa", "Missouri", "North Dakota", "South Dakota",
                   "Nebraska", "Kansas") ~ "Midwest",
      State %in% c("Delaware", "Maryland", "District of Columbia", "Virginia",
                   "West Virginia", "North Carolina", "South Carolina", "Georgia",
                   "Florida", "Kentucky", "Tennessee", "Alabama", "Mississippi",
                   "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
      State %in% c("Montana", "Idaho", "Wyoming", "Colorado", "New Mexico",
                   "Arizona", "Utah", "Nevada", "Washington", "Oregon", "California",
                   "Alaska", "Hawaii") ~ "West"
    )
  )


head(state_only)

ggplot(data = state_only, mapping = aes(x = Year, y = `Age-adjusted Death Rate`)) + 
  geom_line() + 
  facet_wrap(~ State) +
  labs(title = "State Age Adjusted Heart Disease Rate Over Time")


# regional level
regional_summary <- state_only |>
  group_by(Region, Year) |>
  summarise(
    Rate = mean(`Age-adjusted Death Rate`)
  )


ggplot(data = regional_summary, mapping = aes(x = Year, y = Rate, color = Region)) + 
  geom_line(linewidth = 1) +
  labs(title = "Regional Age Adjusted Heart Disease Rate Over Time")

ggplot(data = state_only, mapping = aes(x = Region, y = `Age-adjusted Death Rate`, color = Region)) +
  geom_boxplot(staplewidth = 0.5) +
  labs(title = "Age-Adjusted Heart Disease Mortality by Region")

```


# Statistical Testing

- How does mortality vary across states, and are there regional/geographical patterns over time?

- one way anova limitations:
  - has to be done using a single year, not all years
  - anova assumes independent groups
  - using all years violate independence

- two way anova


- notes:
  - https://www.geeksforgeeks.org/data-science/analysis-of-variance-anova/
  - cook's distance
    - How does a single observation (data pt) influece the entire model?
    - If a data pt is removed, does the model's prediction change a lot?
  
```{r}
# two way anova testing
# * (interaction term) bc each state has their own slope, if + was used then slope for Year is the same across all states
two_way <- aov(`Age-adjusted Death Rate` ~ State * Year, data = state_only)
summary(two_way)

# check assumptions: homoscedasticity, normalit of residuals, linearity, and influential pts
par(mfrow = c(2,2))
plot(two_way)

D <- cooks.distance(two_way)
sum(D > 1)


```



# Predictive Modeling

- Can past trends in age adjusted heart disease death rates be used to predict future state level rates?

- notes:
  - simple linear regression wont work bc we need year and state 
    - assumes all states share the same mean and trend over time
  
  - multiple linear regression year + state
    - treats state as a categorical predictor (fixed state effects)
    - each state getes it's own intercept (baseline level)
    - BUT it assumes same slope for all states
    
  - multiple linear regression year * state 
    - different slope per state

```{r}
# simple linear regression
m1 <- lm(`Age-adjusted Death Rate` ~ Year, data = state_only)
#summary(m1)


# multiple regression +
m2 <- m_additive <- lm(`Age-adjusted Death Rate` ~ Year + State, data = state_only)
#summary(m2)

# multiple regression *
m3 <- lm(`Age-adjusted Death Rate` ~ Year * State, data = state_only)
#summary(m3)

# mixed model?

#mixed1 <- lmer(`Age-adjusted Death Rate` ~ Year + (1 | State), data = state_only)
#summary(mixed1)

mixed2 <- lmer(`Age-adjusted Death Rate` ~ Year + (Year | State), data = state_only)
#summary(mixed2)


# plot comparison

state_only <- state_only |>
  mutate(
    fitted_m1 = fitted(m1),
    fitted_m2 = fitted(m2),
    fitted_m3 = fitted(m3),
    #fitted_mixed1 = fitted(mixed1),
    fitted_mixed2 = fitted(mixed2)
  )

head(state_only)


ggplot(data = state_only, mapping = aes(x = Year)) +
  
  # actual data 
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  # linear
  geom_line(mapping = aes(y = fitted_m1, color = "Simple Linear")) +
  
  # year + state
  geom_line(mapping = aes(y = fitted_m2, color = "Fitted: Year + State"), linetype = "dashed") +
  
  # year * state
  geom_line(mapping = aes(y = fitted_m3, color = "Fitted: Year * State"), linetype = "dotted") +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(
    title = "Model Comparison: Simple Linear vs. Year + State vs. Year x State Models",
    y = "Age-Adjusted Heart Disease Death Rate"
  ) +
  
  scale_color_manual(values = c(
    "Actual" = "black",
    "Simple Linear" = "orange",
    "Fitted: Year + State" = "blue",
    "Fitted: Year * State" = "red"
  ))


ggplot(data = state_only, mapping = aes(x = Year)) +
  # actual data
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  # m3
  geom_line(aes(y = fitted_m3, color = "Fixed: Year * State"), linetype = "dashed") +
  
  # mixed1
  #geom_line(aes(y = fitted_mixed1, color = "Mixed: (1 | State)"), linetype = "dashed") +
  
  # mixed2
  geom_line(mapping = aes(y = fitted_mixed2, color = "Mixed: (Year | State)"), linetype = "dotted") +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(
    title = "Actual vs Fixed-Effects and Mixed-Effects Models by State",
    y = "Age-Adjusted Heart Disease Death Rate",
    color = ""
  ) +
  
  scale_color_manual(values = c(
    "Actual" = "black",
    "Fixed: Year * State" = "blue",
    #"Mixed: (1 | State)"  = "darkgreen",
    "Mixed: (Year | State)" = "red"
  ))

```


- Assumption checks for m1, m2,m3, 

```{r}
par(mfrow=c(2,2))
plot(m1)
plot(m2)
plot(m3)


```

- Train Models:
  - 80/10/10 does not work bc we cant randomly shuffle years

```{r}
train <- state_only |>
  filter(Year <= 2013)

test <- state_only |>
  filter(Year > 2013)

# train models
m2_train <- lm(`Age-adjusted Death Rate` ~ Year + State, data = train)
m3_train <- lm(`Age-adjusted Death Rate` ~ Year * State, data = train)
mixed1_train <- lmer(`Age-adjusted Death Rate` ~ Year + (1 | State), data = train)
mixed2_train <- lmer(`Age-adjusted Death Rate` ~ Year + (Year | State), data = train)

# predict on test set
test$pred_m2 <- predict(m2_train, newdata = test)
test$pred_m3 <- predict(m3_train, newdata = test)
test$pred_mixed1 <- predict(mixed1_train, newdata = test)
test$pred_mixed2 <- predict(mixed2_train, newdata = test)

ggplot(data = test, mapping = aes(x = Year)) +
  
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  geom_line(mapping = aes(y = pred_m3, color = "Predicted (m3)")) +
  
  geom_line(mapping = aes(y = pred_mixed1, color = "Predicted (mixed m1)")) +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(title = "Actual vs Predicted Heart Disease Mortality (Test Set: 2014–2017)") +
  
  scale_color_manual(values = c("Actual" = "black", "Predicted (m3)" = "blue", "Predicted (mixed m1)" = "red"))


```


# Evaluation and Interpretation

- look at RMSE and MAE

```{r}
# check models accuracies
# RMSE and MAE
rmse_m2 <- rmse(test$`Age-adjusted Death Rate`, test$pred_m2)
mae_m2  <- mae(test$`Age-adjusted Death Rate`, test$pred_m2)

rmse_m3 <- rmse(test$`Age-adjusted Death Rate`, test$pred_m3)
mae_m3  <- mae(test$`Age-adjusted Death Rate`, test$pred_m3)

rmse_mixed1 <- rmse(test$`Age-adjusted Death Rate`, test$pred_mixed1)
mae_mixed1  <- mae(test$`Age-adjusted Death Rate`, test$pred_mixed1)

rmse_mixed2 <- rmse(test$`Age-adjusted Death Rate`, test$pred_mixed2)
mae_mixed2  <- mae(test$`Age-adjusted Death Rate`, test$pred_mixed2)

# accuracies
data.frame(
  Model = c("m2 (Year+State)", "m3 (Year*State)", 
            "mixed1 (1|State)", "mixed2 (Year|State)"),
  RMSE = c(rmse_m2, rmse_m3, rmse_mixed1, rmse_mixed2),
  MAE  = c(mae_m2, mae_m3, mae_mixed1, mae_mixed2)
)

```
Between the two models, the predictions follow a smooth linear decline because the models learn long term state specific trends but it can't capture short term year to year flucatuatons. Based on the RMSE and MAE Year x State (m3) performed the best as it showed the lowest scores, whicbh means allowing each state to have its own slope improves predictive model performance. Mixed effects were a bit less accurate because it tries to make the slopes to be more similar across states. This means that past linear trends alone are not enough for accurate year to year predictions of state level heart disease mortality. 


Random Forest
```{r}
library(randomForest)
set.seed(54)
train$State = factor(train$State)
test$State = factor(test$State)

rm_model = randomForest(`Age-adjusted Death Rate`~Year+State, data = train, ntree = 1000, mtry = 2, importance = T)

print(rm_model)

test$pred_rm = predict(rm_model, newdata=test)
rmse_rfm = rmse(test$`Age-adjusted Death Rate`, test$pred_rm)
mae_rfm = mae(test$`Age-adjusted Death Rate`, test$pred_rm)

```

```{r}
data.frame(
  Model = c("m2 (Year+State)", "m3 (Year*State)", 
            "mixed1 (1|State)", "mixed2 (Year|State)", "RF (Year + State"),
  RMSE = c(rmse_m2, rmse_m3, rmse_mixed1, rmse_mixed2, rmse_rfm),
  MAE  = c(mae_m2, mae_m3, mae_mixed1, mae_mixed2, mae_rfm)
)
```
```{r}
# R^2
y_test = test$`Age-adjusted Death Rate`
y_rmf = test$pred_rm

ss_res = sum((y_test - y_rmf)^2)
ss_tot = sum((y_test-mean(y_test))^2)
R2_rmf = 1 - ss_res/ss_tot

R2_rmf
```
So about 95% of the variation in state-level age-adjusted heart-disease mortality(2014-2017) is explained by the random forest model using only Year and State

```{r}
ggplot(data = test, mapping = aes(x = Year)) +
  
  geom_line(mapping = aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  geom_line(mapping = aes(y = pred_m3, color = "Predicted (m3)")) +
  
  geom_line(mapping = aes(y = pred_mixed1, color = "Predicted (m1)")) +
  
  geom_line(mapping = aes(y = pred_rm, color = "Random Forest Model"))+
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(title = "Actual vs Predicted Heart Disease Mortality and Random Forest (Test Set: 2014–2017)") +
  
  scale_color_manual(values = c("Actual" = "black", "Predicted (m3)" = "blue", "Predicted (mixed m1)" = "red", "Random Forest Model" = "green"))
```


```{r}
#Chapter 8 Diagnostic Tests

#Residulas vs Fitted Values
m3 <- lm(`Age-adjusted Death Rate` ~ State * Year, data = state_only)
par(mfrow=c(2,2))
plot(m3, 1)

#Curvature Test
residualPlot(m3) # gives quadratic test for year

#Additional Test - add a quadratic year term 
m3_quad = update(m3,~.+I(Year^2))

#summary(m3)
#summary(m3_quad)

residualPlot(m3_quad)

#Cook Test
D_m3 <- cooks.distance(m3)
sum(D_m3 > 1)

qqnorm(resid(m3), main = "QQ Plot of Residuals for m3")
qqline(resid(m3), col = "navy")

#Extra - Shapiro-Wilk Test
shapiro.test(resid(m3))
```



#Quantinfying whether states are becoming more similar or different 
```{r}
state_range = state_only |>
  group_by(Year) |>
  summarise(mean_val = mean(`Age-adjusted Death Rate`),
            sd_val = sd(`Age-adjusted Death Rate`),
            iqr_val = IQR(`Age-adjusted Death Rate`))

ggplot(state_range, aes(x = Year, y = sd_val))+
  geom_line()+
  labs(
    title = "Variation Between States in Heart Disease Mortality over time",
    y = " Standard Deviation of Age-adjusted Death Rate"
  )

ggplot(state_range, aes(x = Year, y = iqr_val))+
  geom_line()+
  labs(
    title = "Interquantile Range of State Heart Disease Mortality",
    y = "Interquantile Range of Age-adjusted Death Rate"
  )
```

Both SD and IQR of state heart-disease mortality fell
While overall rates declined nationally, state rates became more similar to each other




```{r}
library(splines)


m_quad <- lm(`Age-adjusted Death Rate` ~ (Year + I(Year^2)) * State,
             data = state_only)

par(mfrow = c(2,2))
plot(m_quad)


m_spline <- lm(`Age-adjusted Death Rate` ~ ns(Year, df = 4) * State,
               data = state_only)

par(mfrow = c(2,2))
plot(m_spline)


# Add fitted values to the data
state_only <- state_only |>
  mutate(
    fitted_spline  = fitted(m_spline)   # spline
  )

# Plot: Actual vs Linear vs Spline
ggplot(data = state_only, aes(x = Year)) +
  
  # actual data 
  geom_line(aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  # spline
  geom_line(aes(y = fitted_spline, color = "Spline: ns(Year, 4) * State"),
            linetype = "dashed") +
  
  facet_wrap(~ State, scales = "free_y") +
  labs(
    title = "Model Comparison: Simple Linear vs Spline by State",
    y = "Age-Adjusted Heart Disease Death Rate",
    color = ""
  ) +
  scale_color_manual(values = c(
    "Actual"                      = "black",
    "Spline: ns(Year, 4) * State" = "purple"
  ))
```

```{r}

set.seed(54)
# train/test spline model

m_spline <- lm(`Age-adjusted Death Rate` ~ ns(Year, 4) * State, data = train)


# spline
test$pred_spline <- predict(m_spline, newdata = test)

rmse_spline  <- rmse(test$`Age-adjusted Death Rate`, test$pred_spline)
mae_spline   <- mae(test$`Age-adjusted Death Rate`, test$pred_spline)

ggplot(data = test, aes(x = Year)) +
  
  # actual data
  geom_line(aes(y = `Age-adjusted Death Rate`, color = "Actual")) +
  
  # spline predicted
  geom_line(aes(y = pred_spline, color = "Spline Model"), linewidth = 1) +
  
  facet_wrap(~ State, scales = "free_y") +
  
  labs(
    title = "Actual vs Spline Model Predictions (Test Set)",
    y = "Age-Adjusted Heart Disease Death Rate",
    color = ""
  ) +
  
  scale_color_manual(values = c( 
    "Actual" = "black",
    "Spline Model" = "purple"
  ))

data.frame(
  Model = c("Spline Model (ns(Year,4)*State)",
            "Random Forest (Year + State)"),
  
  RMSE = c(rmse_spline, rmse_rfm),
  MAE  = c(mae_spline, mae_rfm)
)


```
